Learning CNN filters from user-drawn image makers for coconut-tree image classification

### 2020-06-12 main.tex
new resutls
Updates from Overleaf

A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation in those regions by eliminating negative activations, and aggregate the resulting activations (pooling and flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success. Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and use fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually rely on  weight optimization by backpropagation, which requires a higher number of annotated images as deeper is the model. In FLIM, we eliminate the need for backpropagation by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

In FLIM, the user selects a few training images $I$ to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. We wish to estimate the weights of filters $F\in {\cal F}$, such that the convolution between an image $I$ and $F$ will generate positive activations in regions whose local patterns are detected by $F$. In order to understand this process, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect\left[P_I(p)\right] \cdot  \vect\left[F\right]$, where $P_I(p)$ is a patch with $k\times k$ pixels and $m$ image channels around $p$, $F$ is a filter with $k\times k\times m$ weights, and $\vect\left[.\right]$ is the vectorization operation. A filter $F$ can discriminate a class $i\in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect\left[P_I(p)\right]$ of class $i$ and negative results for patterns from other classes. That is, $\vect\left[F\right]$ plays the role of a normal vector of a hyperplane in $\Re^{k\times k\times m}$ and we expect that ${\cal F}$ will contain filters to detect local patterns from all classes in different positions $p$.

For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect\left[P_I(p)\right]$ from all $c$ classes.  
\begin{eqnarray}
{\cal P} & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p)\in [1,c]} P_I(p)
\label{eq.pset}
\end{eqnarray}
The patch set ${\cal P}$ defines samples from all classes in $\Re^{k\times k\times m}$ and we should expect that a clustering operation will create groups with similar local patterns. Ideally, each group should contain local patterns of a same class. However, especially in the first layers, some classes might share similar patterns and we should at least expect that they will be detected in different regions of the image. The groups must be shifted to the origin of $\Re^{k\times k\times m}$ by subtracting the mean vector of the samples in ${\cal P}$. Optionally, the standard deviation of each feature can be computed and used for feature standardization. We call this operation \emph{patch-based normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$ and we also force norm $1$ to avoid preferences among them. This centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane represented by $\vect\left[F\right]$. Figure~\ref{fig:filter} illustrates how it works. ....

\begin{comment}
\begin{figure}[t]
  \begin{center}
     \includegraphics[width=0.8\linewidth]{figures/filter.png}
  \end{center}
  \caption{Filter learned from clustering.}
  \label{fig:filter}
\end{figure}
\end{comment}

%Let $F$ be a filter with dimension $k \times k$, and let $\pi$ be the patch around pixel $(x,y)$ also with dimension $k \times k$. By the definition of convolution in deep learning, we can interpret this operation for a pixel $(x, y)$ as the dot product between the kernel vector $\vect(F)$ and the vector of a patch $\vect(\pi)$ around this pixel, that is, the convolution in $(x, y)$ is the same as $\vect(F) \cdot \vect(\pi)$. Thus, let $H$ be a hyperplane orthogonal to vector $\vect(F)$. Vectors similar to $\vect(F)$ must fall on the same side of the hyperplane as $\vect(F)$. Therefore, if vector $\vect(\pi)$ is on the same side of $H$ as $\vect(F)$, then the dot product between $\vect(\pi)$ and $\vect(F)$ is positive, and it is negative otherwise. Thus, the dot product between the centroid of a cluster and each of the vectors of that cluster must be positive, which implies that the centroid is a good filter to identify the pattern described by that cluster. Hence, if we throw away the negative values, applying convolution with centroids as filters must highlight regions with a certain pattern that the network designer believes is important.

%Since all patches used to produce the filters are centered around the origin, as well as the images of $\D$, we have that the hyperplanes orthogonal to the filters contain the origin. Therefore, there is no need for bias.

Each convolutional layer is trained individually, one layer at a time. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max pooling operation to aggregate local information. We preserve the convolutional layers' output dimensions so that we can use the same markers used in the previous layer. The new patches are generated from the previous convolutional layer output, and the whole process is repeated to find the filters of the next one.

\begin{figure}[!t]
  \centering
  \subfloat[\label{fig:ex-groups}]
  {\includegraphics[width=.35\linewidth]{figures/example_groups/groups_centered}}
  ~
  \subfloat[\label{fig:ex-groups-after}]{\includegraphics[width=.45\linewidth]{figures/example_groups/groups3d_centered}
  }
  \caption{Example of application of convolution with groups centers as filters' weights: (a) Groups of three classes centered on the origin; (b) Groups after applying convolution and activation operations. }
  \label{fig:filter}
\end{figure}

@@

  \end{tikzpicture}
  \end{center}
  \label{fig:tree}
\end{figure}

To exemplify our method, we create a synthetic dataset of points of two classes. We can see this dataset in Figure \ref{fig:ex-groups}. The points are distributed in two regions, and it is not possible to separate the yellow class linearly from the violet class. To center the points,  we subtracted the mean and divided by the standard deviation. If we look only to violet points, it is clear that they form two clusters. In turn, the yellow dots form a single cluster. We then take the center of each of these three clusters and apply convolution to the set of points using these centers as filters. The result of the convolution after applying the ReLU function can be seen in Figure \ref{fig:ex-groups-after}. For most points after the transformation made by convolution and the ReLU function, only one coordinate was positive, this being the coordinate which value is the internal product with the center of the cluster of which these points are part. These points lie along the three axes of space $\mathbb{R}^3$, each cluster on an axis. The other points are farthest from their cluster center, and the internal product with the center of another cluster is also positive. Cluster centers functioned as good filters to detect similar points, even though they detected points from other clusters.

>>>>

A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation by eliminating negative activations, and aggregate the resulting activations (pooling and flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success. Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and use fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification. The training of a CNN usually rely on  weight optimization by backpropagation, which requires a higher number of annotated images as deeper is the model. In FLIM, we eliminate the need of backpropagation to train convolutional layers by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images $I$ to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. The convolution between an image $I$ and a filter $F\in {\cal F}$ generates positive activations in regions whose local patterns are detected by $F$. We wish then to estimate the weights of $F$ such that those local patterns are characteristic of one given class. For sake of clarity, one must interpret the convolution operation at a pixel $p$ of $I$ as the inner product $\vect\left[P_I(p)\right] \cdot  \vect\left[F\right]$, where $P_I(p)$ is a patch with $k\times k$ pixels and $m$ image channels around $p$, $F$ is a filter with $k\times k\times m$ weights, and $\vect\left[.\right]$ is the vectorization operation. A filter $F$ can discriminate a class $i\in \{1,2,\ldots,c\}$ among $c$ classes when it generates positive results for local patterns $\vect\left[P_I(p)\right]$ of class $i$ and negative results for patterns from other classes. That is, $\vect\left[F\right]$ is the normal vector of a hyperplane in $\Re^{k\times k\times m}$ which detects the patterns of class $i$ at its positive side. We wish to estimate ${\cal F}$ such that its filters will detect patterns from all classes in different positions $p$.

For a problem with $c$ classes where $\lambda(p)=i\in \{1,2,\ldots,c\}$ is the label of a marker pixel $p$ from class $i$, let ${\cal M}_I$ be a set of marker pixels drawn in image $I\in {\cal D}$ and $P_I(p)$ be the respective patch around a pixel $p\in {\cal M}_I$. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect\left[P_I(p)\right]$ from all $c$ classes.  
\begin{eqnarray}
{\cal P} & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p)\in [1,c]} P_I(p)
\label{eq.pset}
\end{eqnarray}
The patch set ${\cal P}$ defines samples from all classes in $\Re^{k\times k\times m}$ and a clustering operation on ${\cal P}$ creates groups with similar local patterns. Ideally, each group should contain patterns of a same class. However, some classes might share similar patterns. We should at least expect that their patterns be detected in different image regions. The groups must be shifted to the origin of the feature space by subtracting the mean vector of the samples in ${\cal P}$. Optionally, the standard deviation of each feature can be computed and used for feature standardization. We call this operation \emph{patch-based normalization}. The centroid of each cluster defines the weights of each filter $F\in {\cal F}$ and we force norm $\|\vect\left[F\right]\|=1$ to avoid preferences among them. The centralization for filter definition is paramount to eliminate activations in regions from other classes, whose patterns fall in the negative side of the hyperplane with normal $\vect\left[F\right]$. Figure~\ref{fig:filter}a illustrates three groups with color-coded samples from two classes in a hypothetical 2D feature space. The path-based normalization followed by convolution with filters obtained as centroids of the three groups creates a new 3D feature space (Figure~\ref{fig:filter}b), in which the classes can be more easily detected by two filters.

\begin{figure}[!ht]
  \centering
  \subfloat[\label{fig:ex-groups}]
  {\includegraphics[width=.35\linewidth]{figures/example_groups/groups_centered}}
  ~
  \subfloat[\label{fig:ex-groups-after}]{\includegraphics[width=.45\linewidth]{figures/example_groups/groups3d_centered}
  }
  \caption{(a) A 2D feature space with color-coded samples from two classes distributed in three groups. (b) After patch-based normalization and convolution with filters obtained from each group, the classes can be more easily detected by two filters in a new 3D feature space.}
  \label{fig:filter}
\end{figure}

Each convolutional layer is trained individually, one layer at a time, to find its filter set ${\cal F}$. The number of filters per layer depends on the clustering technique. We use $K$-means, then the user must specify the number $K=|{\cal F}|$ of filters. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max-pooling operation to aggregate local information. Note that, apart from the initial marker selection, the training process is automatic. We have preserved the convolutional layers' output dimensions, such that we can use the same markers used in the input layer to find the filters of layer $L+1$ from the resulting images obtained at the output of layer $L$. Figure~\ref{fig:tree} shows one example of a simple network projected by FLIM for the experiments of the next section.

@@

  \end{tikzpicture}
  \end{center}
  \caption{Example of a network with a single convolutional layer followed by a classifier, as projected by FLIM and used for the experiments in this work.}
  \label{fig:tree}
\end{figure}
